---
title: "Analyzing Recruiting Rankings and College Football Performance"
author: "Winston Platt"
date: "November 2025"
format: html
---

# Overview

This **simplified showcase** demonstrates an analysis of how college football recruiting rankings relate to on-field performance. The original project was done in partnership with Nate Gordon in Fall 2024, but has since been updated. 


```{r}
library(tidyverse)
library(broom)
library(scales) 
```

# Data Preparation

```{r}
df <- read_csv("filtered_data.csv")

glimpse(df)

df <- df %>%
  mutate(
    recruiting_rank = as.numeric(recruiting_rank),
    total_wins = as.numeric(total_wins),
    talent_rating = as.numeric(talent_rating),
    win_pct = total_wins / total_games
  ) %>%
  filter(!is.na(recruiting_rank) & !is.na(win_pct), recruiting_rank > 0)

```


The plots below highlight the relationship between recruiting rank and win percentage.

```{r}
# Scatterplot with smoothing line (lower rank is better)
ggplot(df, aes(x = recruiting_rank, y = win_pct, label = team)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  scale_x_reverse() + # if lower recruiting rank = better, reverse axis for readability
  labs(
    title = "Recruiting Rank vs Win Percentage",
    subtitle = "Lower recruiting rank indicates higher-rated recruiting class",
    x = "Recruiting Rank (lower is better)",
    y = "Win Percentage"
  ) +
  theme_minimal(base_size = 12)
```

*Interpretation:* Teams with better recruiting ranks (closer to 1) tend to have higher win percentages, though variance increases for mid-ranked teams — suggesting other factors (coaching, player development) matter too - otherwise there would be little parity in sports in general.

```{r}
# Boxplot by conference
# Boxplot by conference
ggplot(df, aes(x = conference, y = win_pct)) +
  geom_boxplot(outlier.size = 1) +
  coord_flip() +
  labs(title = "Win Percentage Distribution by Conference", x = "Conference", y = "Win Percentage") +
  theme_minimal(base_size = 11)

```

# Regression & Key Findings

Below is a brief linear model quantifies the predictive relationship between recruiting rank and win percentage.

```{r}
# Simple linear model (win_pct ~ recruiting_rank)
model <- lm(win_pct ~ recruiting_rank, data = df)
tidy(model, conf.int = TRUE) %>% print()

summary(model)$r.squared
glance <- glance(model)
glance %>% print()

ggplot(df, aes(x = recruiting_rank, y = win_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Relationship Between Recruiting Rank and Win Percentage",
    subtitle = paste("R-squared:", round(summary(model)$r.squared, 3)), 
    x = "Recruiting Rank",
    y = "Win Percentage"
  ) +
  theme_minimal()
```

---

# Reproducibility & Contact

**For a deeper analysis:** visit my GitHub profile to see the complete project. 

**Contact:** Winston Platt — winstonrplatt@gmail.com — https://github.com/winstonrplatt

---
